---
title: "BOF"
author: "DEP"
format: html
editor: visual
---

## Purpose

The purpose of this code is to pre-process data for the BOF JAGS model

### Load all libraries

```{r}
library(tidyverse)
library(sf)
library(dplyr)
#needed to save maps. on new systems, may have to do: webshot::install_phantomjs()
library(webshot) 
library(mapview) 
library(tmap)
library(foreign)
library(viridis)
library(fields)
#library(dclone) # built-in functionality for parallel MCMC with jags via jags.parfit()
#library(R2jags)
#library(rjags)
```

### Set directory

```{r}
curr_dir <- dirname(rstudioapi::getSourceEditorContext()$path)
print(curr_dir)
setwd(curr_dir)
```

### Set years & months

```{r}
begYEAR = 1988
endYEAR = 2020

begMONTH = 8
endMONTH = 9
```

### Check for / get data

Identify file location. If file doesn't exist, and if the *googledrive* package is not installed (such as on 2i2c) then install it and proceed. Finally, download the data from your google drive.

```{r}

file_loc = paste0(curr_dir, "/data/", "Copy of Dan & Kelsey All FUNDY data 05-19-2023.CSV")

# if file does not exist ... ask if google drive is installed
if (!file.exists(file_loc)){  
  # return a logial saying if the package is installed
  is_googledrive_installed <- require("googledrive") 
  
  # install googledrive if necessary
  if (is_googledrive_installed == FALSE){
    install.packages("googledrive")
  }

  #download the data from your google drive
  library("googledrive")
  setwd(paste0(curr_dir, "/data/"))
  drive_download("Copy of Dan & Kelsey All FUNDY data 05-19-2023.CSV")
  setwd(curr_dir)  
}
```

### Read data

```{r}
dat <- read_csv(file = file_loc, 
                col_types = cols(FILEID = col_character(),
                                 EVENTNO = col_double(),
                                 MONTH = col_double(),
                                 DAY = col_double(),
                                 YEAR = col_double(),
                                 GMT = col_double(),
                                 LATITUDE = col_double(),
                                 LONGITUDE = col_double(),
                                 LEGTYPE = col_double(),
                                 LEGSTAGE = col_double(),
                                 ALT = col_double(),
                                 HEADING = col_double(),
                                 WX = col_character(),
                                 CLOUD = col_double(),
                                 VISIBLTY = col_double(),
                                 BEAUFORT = col_double(),
                                 SPECCODE = col_character(),
                                 IDREL = col_double(),
                                 NUMBER = col_double(),
                                 CONFIDNC = col_double())
)
head(dat)
```

### Amend / cull the dataset

```{r}
#drop come columns
dat <- dat %>%
  dplyr::select(-starts_with("BEHAV", ignore.case = FALSE, vars = NULL))
#dplyr::select(-c(WX, CLOUD, HEADING, NUMCALF, ANHEAD, PHOTOS))

#restrict to R/V Nereid
dat <- dat %>%
  filter(PLATFORM == 99)

# discard opportunistic surveys
dat <- dat %>%
  mutate(fileid = str_sub(FILEID, start = 1, end = 1)) %>% 
  filter(fileid == "P" | fileid == "p") %>%
  dplyr::select(-fileid)

#add date column
dat$date_ymd <- as.Date(with(dat,paste(YEAR,MONTH,DAY,sep="-")),"%Y-%m-%d")
dat$date_jday = format(dat$date_ymd,"%j")

# keep only desired years and months
dat <- dat %>%
  filter(YEAR >= begYEAR & YEAR <= endYEAR)
dat <- dat %>%
  filter(MONTH == begMONTH | MONTH == endMONTH)

```

### Generate seasons and append to the dataset

```{r}
source('makeSeasons.R')

# MONTHLY SEASONS
    ssn_beg=rbind(c(8,1), c(9,1))
    ssn_end=rbind(c(8,31),c(9,30))
# 2-MONTH SEASONS
    # ssn_beg=rbind(c(8,1))
    # ssn_end=rbind(c(9,30))
# 2-WEEK SEASONS
#    ssn_beg=rbind(c(8,1), c(8,16), c(9,1), c(9,16))
#    ssn_end=rbind(c(8,15), c(8,31), c(9,15), c(9,30))

# create seasons matrix
season <- makeSeasons(begYEAR,endYEAR,ssn_beg,ssn_end)
ssn_beg_date <- as.Date(paste(season[,1],season[,2],season[,3],sep="-"),"%Y-%m-%d")
ssn_end_date <- as.Date(paste(season[,1],season[,4],season[,5],sep="-"),"%Y-%m-%d")
ssn_no = season$SSN_NO
num_ssn = max(ssn_no)
ssn_no_grpd = season$SSN_GRPD_NO

# insert columns with season and season_grpd
dat$season = NA
dat$season_grpd = NA
for (i in 1:length(ssn_beg_date)){
  I = which(dat$date_ymd >= ssn_beg_date[i] & dat$date_ymd <= ssn_end_date[i])
  dat$season[I] = ssn_no[i]
  dat$season_grpd[I] = ssn_no_grpd[i]
}

head(dat)
```

### Flag on/off effort

```{r}
# normally require sea state 0-3, but sea state will be covariate on detection in this model
dat <- dat %>%
  mutate(on.off.eff = 
           if_else((BEAUFORT <= 6 & 
              #start, continue, end watch while ship not underway
              ((LEGTYPE == 5 & (LEGSTAGE == 1 | LEGSTAGE == 2 | LEGSTAGE == 5)) | 
              #legtype = 6 indicates ship not underway (listening station)
              (LEGTYPE == 6 & (LEGSTAGE == 1 | LEGSTAGE == 2 | LEGSTAGE == 5)) ) & 
              #pre-2020 changes to NARWC Sightings Database, 
                #VISIBLTY >=2 or -1 indicates viz of at least 2 nautical miles. 
                #Negative numbers are no longer used,
                #however this dataset was obtained in 2019 before the change.
              (VISIBLTY >=2 | VISIBLTY == -1) &  
              # if there is a sighting, IDREL must = 3. 
                #If no sightings, then IDREL should be NA
              (IDREL == 3 | is.na(IDREL))), 1, 0)) %>%
  #now replace all NA with 0 because those are off-effort
  mutate(on.off.eff = ifelse(is.na(on.off.eff), 0, on.off.eff)
  )
```

### Reduce size of dataset by removing some columns

```{r}
keep.cols <- c("FILEID", 
               "EVENTNO", 
               "YEAR", "MONTH", "DAY", 
               "BEAUFORT", 
               "LEGTYPE", "LEGSTAGE", 
               "LATITUDE", "LONGITUDE", 
               "SPECCODE", "IDREL", "NUMBER", 
               "date_ymd", "date_jday", 
               "on.off.eff", 
               "season", "season_grpd")
tmpdat <- dat %>%
  dplyr::select(all_of(keep.cols)) #%>%
rm(keep.cols)
```

### Option to write out data

```{r}
#write.csv(dat, file = "dat_with_dist.csv")
#write.csv(tmpdat, file = "tmpdat.csv")
```

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

The `echo: false` option disables the printing of code (only output is displayed).
